Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-07-13 paper_jasa.tex
re-worked parts of section 2

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
Note that for $x=0$ the $p$-value becomes --mathematically-- equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

\noindent Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\begin{tabular}{rrrrrrrrrrrrrrrrrrr}
$K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$ &  $x$ & $p$-value && $K$  & $x$ & $p$-value && $K$  & $x$ & $p$-value\\ \cline{1-3} \cline{5-7}\cline{9-11}\cline{13-15}\cline{17-19}
1 &  1 & 0.0500 && 2 &  1 & 0.0975 && 3 & 1 & 0.1426 && 4 & 1 & 0.1855 && 5 & 1 & 0.2262 \\%\cline{1-3}
&&&& 2 &  2 & 0.0025 && 3 & 2 & 0.0073 && 4 & 2 & 0.0140 && 5 & 2 & 0.0226 \\
&&&& &&&& 3 & 3 & 0.0001 && 4 & 3 & 0.0005 && 5 & 3 & 0.0012 \\%\cline{1-6}
&&&&  &    &        &   &   &        &&& 4 & 4 & $< 0.0001$ && 5 & 4 & $< 0.0001$ 
\end{tabular}
\end{center}
\label{pvalue}
\end{table}

@@

\hh{If there is evidence for individual skills influencing the probability $p_i$, then $X_i \sim Binom_{1, p+i}$ and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition, see \citet{butler93} for details. Estimating $p_i$ will be one of the main concerns for the remainder of the paper.} One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

@@

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0 \mid \theta) 
%      = P(X \ge x_{\alpha}) =
%        \begin{cases} 
%               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true}.\\
%              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
%        \end{cases}
    \end{equation*}
\end{dfn}

\hh{\noindent An important difference between conventional and visual testing  is that lineups always depend on observers -- this brings the number of people evaluating a particular lineup into the actual evaluation of power, i.e. we are assuming, again, that $X$ counts the number of observers (out of $K$) who identify the data plot from the lineup.
Then power is calculated as
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
where $F_{X, \theta}$ is the distribution of $X$ -- this depends on which hypothesis is true:

Under the null hypothesis $X \sim Binom_{K, 1/m}$, leading to a type I error of:
\[
\text{Power}_V(\theta, K)= 1 - Binom_{K, 11/m} (x_\alpha - 1),
\]
where $x_\alpha$ is defined as in (\ref{dfn:test}), as  $P(X \ge x_{\alpha}) < \alpha$.

If the alternative hypothesis is true with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Again, assessing an individual's probability to identify the data plot  is central to the actual calculation .
}
%\noindent Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.

@@

Power is an important consideration in deciding which test to use for solving a problem. %In conventional testing, new test procedures are examined for their power under a variety of scenarios. 
For visual inference, this corresponds to the choice of plot to use. Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. %Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.

>>>>

\noindent with $X$ defined as above, and $x$ is the number of observers selecting the actual data plot. 
\end{dfn}
\noindent Note that for $x=0$ the $p$-value becomes, mathematically, equal to 1. It might make more sense from a practical point of view to think of the $p$-value as
being larger than $P(X \ge 1)$ in this situation. If we increase either $m$ or $K$, we would be able to determine the value at a higher precision.

Table \ref{pvalue} shows $p$-values for different numbers of observers for lineups of size $m = 20$. It can be seen that, if the null hypothesis is true, it is unlikely that multiple observers would pick the actual data plot.

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\ \hline %\cline{1-3}\cline{4-6}\cline{7-9}\cline{10-12}\cline{17-19}
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &           &   &        && 4 & 4 & $< 0.0001$ & 5 & 4 & $< 0.0001$ \\\hline
\end{tabular}
\end{center}
\label{pvalue}
\end{table}

@@

If there is evidence for individual skills influencing the probability $p_i$, then \green{$X_i \sim Binom_{1, p_i}$} and $X$ is a sum of independent Bernoulli random variables with different success rates $p_i$. This makes the distribution of $X$ a Poisson-Binomial by definition (see \citet{butler93} for details). \green{Ways to estimate $p_i$ will be be discussed in the following sections.}
%one of the main concerns for the remainder of the paper. One way we can estimate this is fitting a generalized mixed effect model as described in Section \ref{sec:model}.

@@

%\red{How are $\Theta_0$ and $\Theta^c_0$ defined? - why do we need them? wouldn't it be better to re-formulate power in terms of acceptance region, not in terms of the parameter space? - $1/m$ also doesn't look right for the power - I think that's just looking at one observer.}
%\begin{dfn} \label{dfn:power}
%The \textbf{power} of a visual test $V_{\theta}$ is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
%    \begin{equation*}
%      \text{Power}_V(\theta)= 
%        \begin{cases} 
%              \frac1m & \text{if $\theta \in \Theta_0$,} \\
%              Pr(\text{Reject } H_0) &\text{if $\theta \in \Theta^c_0$.}
%        \end{cases}
%    \end{equation*}
%\end{dfn}
%
%
%\red{Alternative Power definition:}
\begin{dfn} \label{dfn:power}
The \textbf{power} of a visual test, $V_{\theta}$, is defined as the probability to reject the null hypothesis for a given parameter value $\theta$:
    \begin{equation*}
      \text{Power}_V(\theta)= Pr(\text{Reject } H_0 \mid \theta) 
%      = P(X \ge x_{\alpha}) =
%        \begin{cases} 
%               1 - \text{Binom}_{K, 1/m}(x_{\alpha} - 1) & \text{ if } H_0 \text{ is true}.\\
%              1 - F_{X, \theta} (x_{\alpha} - 1) & \text{ if } H_1 \text{ is true}.
%        \end{cases}
    \end{equation*}
\end{dfn}

\noindent \green{An important difference between conventional and visual testing is that lineups will depend on observers evaluation. Thus, the number of observers affects the estimation of power, where $X$ is the number of observers (out of $K$) who identify the actual data plot from the lineup, and power is estimated by
\[
\widehat{Power}_{V} (\theta) = {Power}_{V, K} (\theta) = 1 - F_{X, \theta} (x_{\alpha} - 1),
\] 
Here $F_{X, \theta}$ is the distribution of $X$, and this depends on which hypothesis is true.
Under the null hypothesis, $X \sim Binom_{K, 1/m}$, leading to:
\[
\widehat{Power}_V(\theta, K)= 1 - Binom_{K, 1/m} (x_\alpha - 1),
\]
where $x_\alpha$ is such that $P(X \ge x_{\alpha}) < \alpha$. On the other hand, if the alternative hypothesis is true, with a fixed parameter value $\theta$, we can assume that an individual's probability to identify the data plot depends on the parameter value, and $X_i \sim Binom_{1, p_i(\theta)}$. Assessing an individual's probability to identify the actual data plot will require that an individual evaluate multiple lineups, which may not always be feasible.}

%\noindent Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.

@@

\green{Power is an important consideration in deciding which test to use for solving a problem. %In conventional testing, new test procedures are examined for their power under a variety of scenarios. 
Here we use it to compare the performance of the visual test with the conventional test, but in practice for visual inference it will mostly be important in deciding on the choice of plot to use.} Analysts typically have a choice of plots to make, and a myriad of possible options such as reference grids, for any particular purpose. This is akin to different choices of statistics in conventional hypothesis testing, for example, mean, median, or trimmed mean. One is typically better than another. For two different visual test statistics of the same actual data, one is considered to be better, if $T(y)$ is more easily distinguishable to the observer. Power is typically used to measure this characteristic of a test. %Power is the probability of rejecting $H_0$ when it is not true.  We can assess power therefore both empirically based on experimental data and through theoretical considerations.
